Space is the Place: How Dispersal Shapes Genetic Variation in Continuous Space

### 2019-03-13 ms/spaceness.tex
ms updates

\documentclass[9pt,twocolumn,twoside]{gsajnl}
\usepackage{amsmath}
% Use the documentclass option 'lineno' to view line numbers

\articletype{inv} % article type
% {inv} Investigation 
% {gs} Genomic Selection
% {goi} Genetics of Immunity 
% {gos} Genetics of Sex 
% {mp} Multiparental Populations

@@

\author[$\ast$,1]{C.J. Battey}
\author[$\dagger$,1]{Peter Ralph}
\author[$\ddagger$,2]{Andrew Kern}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants. As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography. Here we implement a forward-time continuous-space model of molecular evolution and use it to study the impacts of limited dispersal in continuous landscapes on population genetic summary statistics, demographic inference, and association studies. Low dispersal slows the geographic spread of ancestry and inflates branch lengths for mid-frequency alleles because of slow coalescence among distant individuals. This inflation can bias demographic inference from the site frequency spectrum towards inferring a recent spike in population sizes. In a GWAS setting spatially correlated phenotypes cause spurious signals of genetic association when principal components analyses fail to capture the relevant spatial scale of phenotypic variation, which can occur under either very limited or very high dispersal. Last we use a machine learning regression approach to evaluate summary statistics as estimators of dispersal distance and find that correlation coefficients between spatial distance and moments of the distribution of pairwise haplotype block lengths retain strong signal even at high dispersal distances. 
\end{abstract}

@@

\correspondingauthoraffiliation{3}{University of Oregon Dept. Biology, Institute for Ecology and Evolution.}
\vspace{-35pt}% Only used for adjusting extra space in the left column of the first page

@@

NOTE: probably cut this paragraph???

This process is analogous to genetic drift, with spatial locations and population densities taking the place of sites and allele frequencies. Because individuals disperse from their natal location, new offspring are likely to be close to their parents. If by chance an individual has no offspring the local population density is then lowered, which decreases the chance of new individuals dispersing into that area. This process compounds over time until only one or two patches of habitat are occupied at any given time. The resulting spatial clustering clearly does not describe spatial dynamics in most real populations, because declining growth rates in regions of high population density are common in both lab ( ) and wild populations ( ).

@@

Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts will cause inflation of test statistics at SNPs with different allele frequencies among populations (CITE). In addition, because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within and among populations can lead to inflated effect size estimates in GWAS of quantitative traits conducted on samples from large geographic regions \citep{Mathieson2012}. For example, imagine a population of flowering plants occurring over a large latitudinal range with populations subject to isolation by distance. Flowering time is strongly dependent on photoperiod and temperature, both of which typically vary with latitude. If we were to conduct a GWAS for flowering time without correcting for population structure we would find that all SNPs with allele frequencies varying clinally with latitude were significantly associated with flowering time, even if flowering time was determined purely by the environment.

A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}. Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Kang2010}. In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype. Recent studies in European and British populations \citep{Kerminen2018,Berg2018,Sohail2018} have found evidence of residual population stratification in effect size estimates from large-scale GWAS employing both linear and mixed model approaches, suggesting that existing structure corrections may be inadequate. For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high. Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections fail.  
\section{Materials and Methods}
\label{sec:materials:methods}

@@

We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1e8$, and recombination rates $1e-9$. A mutation rate of $1e-8 mutations/site/generation$ was then applied to the coalesced tree sequence with msprime. $\sigma$ values were drawn from a uniform distribution bounded by 0.2 and 4. Simulations were run until all extant individuals shared a common ancestor within the simulation (i.e. the tree sequence had coalesced). Coalescent stepping-stone simulations were run on a 5x5 grid with a total $N_{e}$ of 10,000 ($N_{e}$ per population = 400) and 10 haploid samples per population; allowing migration between all adjacent populations. We ran 500 simulations drawing the number of migrants exchanged by adjacent demes from a uniform distribution between 0 and 40.

@@

Because generation time in our model is not fixed and may vary both across simulations we also calculated "effective" generation times for all simulations. To do this we output the full tree sequence of all individuals in every time step for 200 time steps. Starting at the final time step we then calculated the average age of parents in each previous generation, sampling past individuals in proportion to their genetic contribution to the final generation, which can be extracted from the tree sequence. When adding mutations to the tree sequences produced by SLiM, we scaled the mutation rate by the generation time because msprime's "mutate()" function assumes that time is measured in units of generations. So to achieve a $1e-8 mutations/site/generation$ mutation rate for a simulation in which the generation time was 5 time steps, we entered $1e-8/5$ for the mutation rate in msprime.mutate(). We then verified that this procedure produced the correct number of mutations by comparing a subset simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime, finding that they were nearly identical in all cases.    
\subsection{Summary Statistics}
We calculated a set of 20 summary statistics (Supplementary Table 1) from 60 diploid individuals sampled randomly from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. Statistics included common single-population summaries such as mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.

@@

\subsection{Demographic Modeling}
We fit single-population demographic models to the site frequency spectra of 60 randomly sampled individuals from each simulation with the program Stairwayplot \citep{Liu2015}. Site frequency spectra used for input data were calculated in scikit-allel. 100 bootstrap replicates were generated for each simulation by resampling genotype matrices with replacement over sites and recalculating the site frequency spectra. Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of Ne per time unit across bootstrap replicates was used to represent the output of each simulation. The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of phenotypic variation for 1000 randomly sampled final-generation individuals in each SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 1\% were excluded from the analysis. Phenotype values were meant to roughly reflect the distribution of height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.

@@

\section{Results}
\subsection{Runtime and Memory Requirements}
\subsection{Summary Statistics}
\subsection{Demographic Modeling}
\subsection{GWAS}

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{gwas_summary_sigma225.png}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on GWAS. Colors represent different phenotype distributions as indicated in the figure legend. A: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates. B: Example Manhattan plots for a single simulation with varying phenotype distributions. Points in black are significant at an estimated FDR cutoff of 0.05, while the dotted line shows the Bonferroni-corrrected \textit{p} value of 0.05. In practice most human GWAS studies use a significance cutoff between these points. C and D show the corresponding spatial and principal component positions, colored by phenotype.}
\label{fig:spectrum}
\end{figure*}

@@

>>>>

\documentclass[9pt,twocolumn,twoside,lineno]{gsajnl}
\usepackage{amsmath}
% Use the documentclass option 'lineno' to view line numbers

\articletype{preprint} % article type
% {inv} Investigation 
% {gs} Genomic Selection
% {goi} Genetics of Immunity 
% {gos} Genetics of Sex 
% {mp} Multiparental Populations

@@

\author[$\ast$,1]{C.J. Battey}
\author[$\ast$]{Peter Ralph}
\author[$\ast$]{Andrew Kern}

@@

\begin{abstract}
Individuals exist in continuous space, but standard models in population genetics are based on discrete randomly-mating populations exchanging migrants. As the availability of population-level, whole-genome data allows inference of increasingly fine-scale patterns of ancestry in many species, models incorporating realistic demographic and spatial processes are needed to accurately describe spatial structure and control for its impacts on analyses of selection and demography. Here we implement a forward-time simulation of molecular evolution in continuous space and use it to study the impacts of dispersal and density-dependent competition on population genetic summary statistics, demographic inference, and association studies. Low dispersal slows the geographic spread of ancestry and increases branch lengths corresponding to mid-frequency alleles because of slow coalescence among distant groups of individuals. As a result demographic inference from the site frequency spectrum can be biased towards inferring a spike in past population sizes. Spatially correlated phenotypes cause spurious signals of genetic association when principal components analyses fail to capture the relevant spatial scale of phenotypic variation, which can occur under either very limited or very high dispersal. Last we use a machine learning regression approach to evaluate summary statistics as estimators of dispersal distance and find that correlation coefficients between spatial distance and moments of the distribution of pairwise haplotype block lengths retain strong signal even at high dispersal distances. 
\end{abstract}

@@

\correspondingauthoraffiliation{1}{301 Pacific Hall, University of Oregon Dept. Biology, Institute for Ecology and Evolution. cbattey2@uoregon.edu.}
\vspace{-35pt}% Only used for adjusting extra space in the left column of the first page

@@

@@

Last, we use our model to study the impacts of isolation by distance in continuous space on genome-wide association studies (GWAS). Variation in ancestry proportions between case and control cohorts is expected to cause inflation of test statistics at SNPs with different allele frequencies among populations (CITE). In addition, because most phenotypes are influenced by the environment and environmental factors are often spatially correlated, subtle spatial structure within and among populations can lead to inflated effect size estimates in GWAS of quantitative traits conducted on samples from large geographic regions, and this effect is particularly prominent in low-frequency alleles \citep{Mathieson2012}. For example, imagine a population of flowering plants occurring over a large latitudinal range with populations subject to isolation by distance. Flowering time is strongly dependent on photoperiod and temperature, both of which typically vary with latitude. If we were to conduct a GWAS for flowering time without correcting for population structure we would find that all SNPs with allele frequencies varying clinally with latitude were significantly associated with flowering time, even if flowering time was determined purely by the environment.

A common method of correcting for population structure in GWAS is to run a principal components analysis on the genotype matrix and include PC coordinates as covariates in a linear regression for each SNP \citep{Price2006}. Mixed model GWAS employs a different statistical framework but takes a similar approach by incorporating a kinship matrix describing pairwise genetic relatedness among samples as a random effect in a linear mixed model \citep{Yu2006,Kang2008}. In both approaches the key question is whether or not the ancestry components of the model adequately describe "background" levels of genetic variation in the sample relative to any confounding effects of ancestry and phenotype. Recent studies in European and British populations \citep{Kerminen2018,Berg2018,Sohail2018} have found evidence of residual population stratification in effect size estimates from large-scale GWAS employing both linear and mixed model approaches, suggesting that existing structure corrections may be inadequate. For traits controlled by a few large effect loci this confounding effect is likely to be relatively small, but for analyses of complex traits with thousands of putatively associated SNPs the potential for confounding is high. Here we simulate a range of spatially correlated phenotypes for simulated individuals and seek to identify conditions under which existing GWAS structure corrections fail.  
\section{Materials and Methods}
\label{sec:materials:methods}

@@

We ran 400 simulations for the spatial and random-mating SLiM models on a square landscape of size 50 with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), average lifetime $L=4$, genome size $1\times10^{8}$, recombination rate $1\times10^{-9}$, and drawing $/sigma$ values from a uniform distribution bounded by 0.2 and 4. To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM and later applied mutations to the coalesced tree sequence with msprime's 'mutate( )' function. Because msprime assumes that timesteps are in units of generations, we divided the per-generation rate of $1\times10^{-8} mut/site/gen$ by the generation time estimated for each value of $\sigma$ (see 'Genealogical Parameters' below) to convert the rate to units of mutations/site/timestep. We then verified that this procedure produced the correct number of mutations by comparing a subset of simulations with SLiM-generated mutations (which are applied only at mating events) with mutations added by msprime, finding that they were nearly identical in all cases. Simulations were run until all extant individuals shared a common ancestor within the simulation (i.e. the tree sequence had coalesced).

Coalescent stepping-stone simulations were on a 5x5 grid with 10 haploid samples per deme and migration between adjacent populations. To facilitate comparison across simulation types we ran one coalescent simulation for every SLiM parameter set. Because $\sigma$ is approximately equal to the mean distance between parent and offspring, we approximate the relationship between $\sigma$ and the migration rate in a grid simulation as $\sigma=4md$, where $m$ is the migration rate and $d$ is the ratio of landscape widths in continuous space vs coalescent simulations (here, 10). That is, an individual in an interior population has a $4m$ probability of being a migrant, and if it is a migrant will move $d$ units on the map. We set the total $N_{e}$ equal to the census size of each SLiM simulation divided by the variance in reproductive output (see below).

@@

Next we estimated the variance in reproductive output and the generation time of individuals in our simulations. To do this we output the full tree sequence of all individuals for 200 time steps. Starting at the final time step we then calculated the average age of parents in each previous generation as an estimate of generation time, sampling past individuals in proportion to their genetic contribution to the final generation. To estimate variance in reproductive output we calculated the variance in number of offspring for all individuals born in timesteps 50-150 (to ensure that the simulation captured the full lifespan of the sampled individuals).

\subsection{Summary Statistics}
We calculated a set of 20 summary statistics (Supplementary Table 1) from 60 diploid individuals sampled randomly from the final generation of each simulation using the python package scikit-allel \citep{Miles2017}. Statistics included common single-population summaries such as mean pairwise divergence ($\pi$), inbreeding coefficient ($F_{is}$), and Tajima's D, as well as the classic isolation-by-distance regression of genetic distance ($D_{xy}$) against the logarithm of geographic distances \citep{Rousset1997}, which we summarized as the correlation coefficient between $log10(spatial distance)$ and the proportion of identical base pairs among individuals.

@@

\subsection{Demographic Modeling}
We fit single-population demographic models to the site frequency spectra of 60 randomly sampled individuals from each simulation with the program Stairwayplot \citep{Liu2015}. Site frequency spectra used for input data were calculated in scikit-allel. 100 bootstrap replicates were generated for each simulation by resampling over sites. Models were then fit across all bootstrap replicates using default settings in Stairwayplot and the median estimate of $Ne$ per generation across bootstrap replicates was used to represent the output of each simulation. The resulting inferred population size histories were plotted with line colors scaled to $\sigma$ to visualize how dispersal scaling impacts estimates of demographic history.

\subsection{Association Studies}
To assess the degree to which spatial structure confounds GWAS we simulated four types of phenotypic variation for 1000 randomly sampled final-generation individuals in each SLiM simulation and conducted a linear-regression GWAS with PC covariates in PLINK \citep{PURCELL2007}. SNPs with a minor allele frequency less than 1\% were excluded from this analysis. Phenotype values were meant to roughly reflect the distribution of height across Europe, which has recently been found to be confounded with population structure in large scale GWAS \citep{Berg2018,Sohail2018}. Conceptually our approach is similar to that taken in \citep{Mathieson2012}, though here we model continuous spatial variation and focus on genome-wide effects rather than low-frequency alleles specifically.

@@

\section{Results}

\subsection{Runtime and Memory Requirements}

\subsection{Summary Statistics}

\subsection{Demographic Modeling}

\subsection{GWAS}
Spatially correlated phenotypes in populations with low dispersal distances cause uncorrected association analyses to report significant associations at most genome-wide SNPs (Figure 1). No inflation was observed for the "normal" phenotype distribution which is random with respect to space, demonstrating that this effect is caused by the interaction of isolation by distance and spatially correlated phenotypes. The deflation in $p$ values for spatial phenotypes is particularly pronounced for the corner and clinal distributions, likely because in these cases our sample captures more phenotypic variation than in the "patchy" distribution. When PCA positions are included as covariates the vast majority of SNPs no longer pass a 5\% FDR significance threshold, but up to 1000 genome-wide significant hits (out of an average of 150,000 SNPs) were observed even in relatively high dispersal for the corner and clinal distributions.

Notably, many SNPs inferred as significant are in close proximity along the genome and likely reflect haplotypes that became common in one region of the landscape (Figure 1, supplementary figures X) [note: add example manhattan plots for s=0.3 and s=3 to fig, supp figs for manhattan plots across a larger range of sigma]. Thus the assumption that patterns of linkage disequilibrium are uncorrelated with allele frequency differentiation, which is the basis of LD score regression \citep{Bulik-Sullivan2015}, may not hold for these simulations. ###TODO revisit this and

In the case of clinal phenotypes, PCA covariates perform very well in correcting for population stratification at low dispersal distances but the correction breaks down above neighborhood sizes of 250. This pattern appears to reflect a loss of descriptive power in the first ten axes of the PCA, which describe less of the total variation in the data as neighborhood size increases (Figure 1 PANEL TO BE ADDED). Essentially, PCA loses descriptive power faster than the breakdown in correlations between allele frequencies and spatial position that our GWAS is picking up on, so eventually the statistical correction breaks down.

Examining quantile-quantile plots shows that $p$ value deflation is most pronounced for the corner phenotype distribution and is particularly strong for simulations with neighborhood sizes of less than 20. These results are similar to the observation of high inflation for spatially discrete phenotypes in \citep{Mathieson2012}. Notably here we find that the deflation in $p$ values is not limited to low-frequency alleles, and is not entirely controlled by including PCA covariates. Quantile-quantile plots of clinal and nonspatial phenotypes also show evidence of $p$ value $inflation$ (i.e. points below the 1:1 line), which is observed only when PCA covariates are included in the analysis (see Supplementary Figure X for q-q plots of uncorrected GWAS results). This likely reflects the loss of power in GWAS when both phenotype and ancestry are correlated with space, which creates a partially confounded study design.

@@

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{gwas_summary_nsig_qqplots.pdf}
\caption{Impacts of spatially correlated phenotypes and isolation by distance on GWAS. Colors represent different phenotype distributions as indicated in the figure legend. A: numbers of significant SNPs after FDR correction for linear-model GWAS conducted with (top) or without (bottom) PC covariates. B: Quantile-quantile plots of observed versus $p$ values relative to those expected from a uniform distribution assuming no effect on phenotypes. The dotted lines show the 95\% confidence region assuming binomial sampling and an alpha level of 0.05. Points above the 1:1 line reflect inflated -log10(p).}
\label{fig:spectrum}
\end{figure*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
